---
title: "Remote sensing evidence for the 2020 harmful algal bloom in Kamchatka"
subtitle: "A remote sensing study in the reproducible research framework"
date: July 17, 2022
author: Ivan Dubrovin
format:
  revealjs:
    logo: Skoltech_Logo_Black.png
    theme: [simple, theme.scss]
---

```{r}
#| include: false
#| label: setup
knitr::opts_knit$set(root.dir = here::here())
library(targets)
```

## Two parts of the presentation


# Part 1: The study

## Background

What happened

## Region of interest

## Data

## Methods

$$
\Delta_r \text{Chl-a} = \frac{\text{Chl-a} - \text{Chl-a}^{\text{ref}}}{\text{Chl-a}^{\text{ref}}}.
$$

## Chlorophyll anomaly evolution

# Part 2: The approach

## Literate programming

Literate programming is a programming paradigm introduced by Donald Knuth in which a computer program is given an explanation of its logic in a natural language, such as English, interspersed with snippets of macros and traditional source code, from which compilable source code can be generated.

Any change is automatically included into the output: no copy-pasting of images, no chance to forget to update the final result.

## Constituents

* R Markdown
* Quarto
* Python
* R

## Python and R

Very easy to use together

### Python
- Wrapper around SNAP GPT
- `lxml`

### R
- Processing and visualization of spatial outputs
- `sf`, `terra`, `tmap`, `targets`, `knitr`

## R Markdown and Quarto

## Target network

```{r}
#| echo: false
tar_visnetwork(targets_only = TRUE)
```

::: footer
How to automate the data retrieval?
:::

## How to recreate the results

Run the pipeline.

```r
targets::tar_make()
```

Compile the paper pdf.

```r
knitr::knit()
```

Generate the presentation html.

```shell
quarto render
```

::: footer
Assuming you got the archived Sentinel-3 data.
:::
